---
title: "Final Assignment: Roman Emperors"
author: "Rebeka Zsoldos"
output: html_document
editor_options:
  chunk_output_type: console
---
## Background

According to wikipedia:

The Roman emperor was the ruler of the Roman Empire during the imperial period (starting in 27 BC). The emperors used a variety of different titles throughout history. Often when a given Roman is described as becoming "emperor" in English, it reflects his taking of the title Augustus or Caesar. Another title often used was imperator, originally a military honorific. Early emperors also used the title Princeps Civitatis ('first citizen'). Emperors frequently amassed republican titles, notably princeps senatus, consul, and pontifex maximus.

The legitimacy of an emperor's rule depended on his control of the army and recognition by the Senate; an emperor would normally be proclaimed by his troops, or invested with imperial titles by the Senate, or both. The first emperors reigned alone; later emperors would sometimes rule with co-emperors and divide administration of the empire between them.

The Romans considered the office of emperor to be distinct from that of a king. The first emperor, Augustus, resolutely refused recognition as a monarch.[1] For the first three hundred years of Roman emperors, from Augustus until Diocletian, efforts were made to portray the emperors as leaders of the republic, fearing any association with the kings of Rome prior to the Republic.

From Diocletian, whose tetrarchic reforms also divided the position into one emperor in the West and one in the East, until the end of the Empire, emperors ruled in an openly monarchic style[2] and did not preserve the nominal principle of a republic, but the contrast with "kings" was maintained: although the imperial succession was generally hereditary, it was only hereditary if there was a suitable candidate acceptable to the army and the bureaucracy,[3] so the principle of automatic inheritance was not adopted. Elements of the republican institutional framework (senate, consuls, and magistrates) were preserved even after the end of the Western Empire. (https://en.wikipedia.org/wiki/Roman_emperor)

```{r setup, include=FALSE}
library(dplyr)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(ggfortify)
library(viridis)
library(glmm)
library(broom)
library(sjPlot)
library(car)

theme_set(theme_light())
```

## Importing Data

```{r}
#read data
emperors_raw <- readr::read_csv(url("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-08-13/emperors.csv"), na = c("NA", "Unknown"))

```

## Cleaning the Data


``` {r}
emperors <- emperors_raw %>%
  mutate(
    birth_cty = as.factor(birth_cty),
    birth_prv = as.factor(birth_prv),
    #relevel the factors
    rise = fct_relevel(rise, 
                            "Seized Power", "Appointment by Army", "Appointment by Praetorian Guard", "Purchase",
                            "Election", "Appointment by Senate", "Appointment by Emperor", "Birthright"),
     #relevel the cause of death
    cause = fct_relevel(cause, 
                             "Natural Causes", "Assassination", "Execution",  "Suicide", "Died in Battle",
                              "Captivity"),
    killer = as.factor(killer),
    dynasty = as.factor(dynasty),
    era = as.factor(era)
           ) %>%
  select(-name_full, -notes, - verif_who) %>%
#As the dataset contains BC years so it is necessary to recode them for further computations
#Firstly set the dates to years only
  mutate(
    birth_year = as.numeric(format(birth, '%Y')),
    death_year = as.numeric(format(death, '%Y')),
    reign_start_year = as.numeric(format(reign_start, '%Y')),
    reign_end_year = as.numeric(format(reign_end, '%Y'))) %>%
#Recoding BC years as negative numbers and AC years as positive numbers
#If the birth year or the starting year of the reign is higher then the year of the death/the end of reign, then that year must be BC
  mutate(
    birth_year = ifelse(birth_year > death_year, (birth_year * -1), (birth_year * 1)),
    reign_start_year = ifelse(reign_start_year > reign_end_year, (reign_start_year * -1), (reign_start_year * 1)),
    age_reign_start = reign_start_year - birth_year,
    age_reign_end = reign_end_year - birth_year,
#The length of the reign in years
    reign_long_year = reign_end_year - reign_start_year,
#Which century did the emperor rule?
    century = reign_end_year %/% 100 + 1)
```


## EDA

### How rulers came to power and lost it

``` {r}
#Putting the emperors on a timeline, based on the fact how they rose to power and which dynasty they did belong
rise_emperors <- ggplot(emperors, aes(reign_start_year, rise, color = dynasty, shape = era)) +
  geom_point(alpha = 0.5) +
  scale_x_continuous(breaks = c(-26, 305, 395),
                   labels = c("26 B.C.", "305 A.D.", "395 A.D.")) +
  xlab("The year when their ruling started") +
  ylab("The cause to rising to power") +
  labs(title = "The Rise of the Emperors") +
  scale_color_viridis_d()

rise_emperors
  
#Putting the emperors on a timeline, based on the fact how they fell out of power and which dynasty they did belong
fall_emperors <- emperors %>%
  ggplot(aes(reign_end_year, cause, color = dynasty, shape = era)) +
  geom_point(alpha = 0.5) +
  scale_x_continuous(breaks = c(14, 305, 395),
                   labels = c("14 A.D.", "305 A.D.", "395 A.D.")) +
  xlab("The year when their ruling ended") +
  ylab("The cause to fall out of power") +
  labs(title = "The Fall of the Emperors") +
  scale_color_viridis_d()

fall_emperors

```

The emperors came to power mostly by birthright and died of natural causes or were the victims of assassination.

### How long the reign of the emperors lasted and how it ended

``` {r}

#Exploring the frequency of the lengths of the reigns
reign_length_emperors <- emperors %>%
  ggplot(aes(reign_long_year, fill = cause)) +
  geom_histogram() +
  theme(legend.position = "bottom") +
  xlab("The rise of the emperors") +
  ylab(NULL) +
  labs(title = "The length of the reign of each emperor") +
  scale_fill_viridis(discrete = TRUE, na.value = "grey", name = "The cause of death")

reign_length_emperors

#Exploring how the emperors died categorized by their rise to power
rise_and_fall2 <- emperors %>%
  ggplot(aes(rise, fill = cause)) +
  geom_bar(width = 0.3) +
  coord_flip() +
  theme(legend.position = "bottom") +
  xlab("The rise of the emperors") +
  ylab(NULL) +
  labs(title = "The Rise and Fall of the Roman Emperors") +
  scale_fill_viridis(discrete = TRUE, na.value = "grey", name = "The cause of death")

rise_and_fall2

```

Many emperors ruled for only one or two years, and those who came to power by birthright may also have been assassinated.

### How old were the emperors when they came to power?

``` {r}
#Exploring the frequency of the ages of the emperors at the beginning of their reign
age_rise_emperors <- emperors %>%
  ggplot(aes(age_reign_start, fill = rise)) +
  geom_histogram() +
  theme(legend.position = "bottom") +
  xlab("Age in years") +
  ylab(NULL) +
  labs(title = "The age of the emperors at the beginning of their reign") +
  scale_fill_viridis(discrete = TRUE, na.value = "grey", name = "The cause of rise")

age_rise_emperors

#calculate the mean when the emperors rose to power
mean_age_reign_start <- mean(emperors$age_reign_start, na.rm = TRUE)

#calculate SD
sd_age_reign_start <- sd(emperors$age_reign_start, na.rm = TRUE)


#Exploring how the age of the emperors at the beginning of their reign relates to the duration of their reign
age_and_long <- emperors %>%
  ggplot(aes(age_reign_start, reign_long_year)) +
  geom_point(alpha = 0.2, color = "red") +
  geom_text(aes(label = name), size= 3, check_overlap = TRUE) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  xlab("The age of the emperor when rising to power") +
  ylab("Duration of the reign") +
  scale_color_viridis(discrete = TRUE, na.value = "grey")

age_and_long

```

Average age at which rulers came to power was `r mean_age_reign_start` years (SD: `r sd_age_reign_start`). Those who came to the throne when they were old ruled for shorter periods.

### How old the emperor were when they lost their power?

```{r}

#Exploring the frequency of the ages of the emperors at the beginning of their reign
age_death_emperors <- emperors %>%
  ggplot(aes(age_reign_end, fill = cause)) +
  geom_histogram() +
  theme(legend.position = "bottom") +
  xlab("Age in years") +
  ylab(NULL) +
  labs(title = "The age of the emperors at the end of their reign") +
  scale_fill_viridis(discrete = TRUE, na.value = "grey", name = "The cause of death")

age_death_emperors

#calculate the mean when the emperors fell out of power
mean_age_reign_end <- mean(emperors$age_reign_end, na.rm = TRUE)

#calculate SD
sd_age_reign_end <- sd(emperors$age_reign_end, na.rm = TRUE)

#Exploring how the age of the emperors at the end of their reign relates to the duration of their reign
death_and_long <- emperors %>%
  ggplot(aes(age_reign_end, reign_long_year)) +
  geom_point(alpha = 0.2, color = "red") +
  geom_text(aes(label = name), size= 3, check_overlap = TRUE) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  xlab("The age of the emperor when his reing ended") +
  ylab("Duration of the reign") +
  scale_color_viridis(discrete = TRUE, na.value = "grey")

death_and_long
```

Average age at which rulers came to power was `r mean_age_reign_end` years (SD: `r sd_age_reign_end`).

## Recoding the data to prepare it for the predictive models

```{r}

emperors_principate <- emperors %>%
#Filtering the data for the Principate era only as in the Dominate era there were 4 rulers at once
  filter(era == "Principate") %>%
  dplyr:: select(-birth, -death, -birth_cty, -birth_prv, -reign_start, -reign_end, -era) %>%
  mutate(
#recode cause into three categorical level
#Making one category form Captivity and Died in Battle: Battle
#Making one category from Assassination, Suicide, Execution: Forced  
    cause_recoded = recode_factor(cause, "Died in Battle" = "Battle", "Natural Causes" = "Natural Causes", .default = "Forced"),
#Recoding the factor killer
#Making one category from Aneurism and Heart Failure: Health
#Making one category from Lightning, Fire: Natural disaster
    killer_recoded = recode_factor(killer, "Aneurism" = "Health", "Heart Failure" = "Health", "Disease" = "Health",
                            "Lightning" = "Natural   Phenomenon", "Fire" = "Natural Phenomenon", "Wife" = "Other Opponents",
                            "Court Officials" = "Other Opponents", "Usurper" = "Other Opponents"),
#Recoding the factor rise
#Creating to categories: Birthright, Appointed
    rise_recoded = recode_factor(rise, "Birthright" = "Birthright", .default = "Other"),
#Making dummy variables for dynasties
    d_Gordian = recode_factor(dynasty, "Gordian" = 1, .default = 0),
    d_Julio_Claudian = recode_factor(dynasty, "Julio-Claudian" = 1, .default = 0),
    d_Severan = recode_factor(dynasty, "Severan" = 1, .default = 0),
    d_Nerva_Antonine = recode_factor(dynasty, "Nerva-Antonine" = 1, .default = 0),
#Making dummy variables from the cause_recoded for the models
    died_by_force = recode_factor(cause_recoded, "Forced" = 1, "Battle" = 0, "Natural Causes" = 0),
    died_by_battle = recode_factor(cause_recoded, "Forced" = 0, "Battle" = 1, "Natural Causes" = 0),
    died_naturally = recode_factor(cause_recoded, "Forced" = 0, "Battle" = 0, "Natural Causes" = 1)
    ) %>%
  dplyr:: select(index, name, rise_recoded, cause_recoded, everything(), -cause, -killer, - birth_year, -death_year, -rise, -dynasty, -killer_recoded) %>%
#Removing missing data  
  na.omit()

```

## Predicting the age of the emperors at the end of their reign

```{r}
#checking the normality of the dependent variable
shapiro.test(emperors_principate$age_reign_end)

#MODEL1
model_age_end1 <- lm(age_reign_end ~ rise_recoded + century, data = emperors_principate)
summary(model_age_end1)

#visualizing the fit of MODEL1
autoplot(model_age_end1, which = 1:3)

#Homoskedaticity of MODEL1
student_residuals = rstudent(model_age_end1)
fitted = model_age_end1$fitted
plot(fitted, student_residuals, pch = 19, frame.plot = FALSE, col = "darkgrey")
abline(h=0, lwd=2, col = "red")
ncvTest(model_age_end1)
spreadLevelPlot(model_age_end1)

#MODEL2
model_age_end2 <- lm(age_reign_end ~ rise_recoded + age_reign_start, data = emperors_principate)
summary(model_age_end2)

#visualizing the fit of MODEL2
autoplot(model_age_end2, which = 1:3)

#Homoskedaticity of MODEL2
student_residuals = rstudent(model_age_end2)
fitted = model_age_end2$fitted
plot(fitted, student_residuals, pch = 19, frame.plot = FALSE, col = "darkgrey")
abline(h=0, lwd=2, col = "red")

ncvTest(model_age_end2)
spreadLevelPlot(model_age_end2)

#table output for the models
tab_model(model_age_end1, model_age_end2, show.est = TRUE, show.fstat = TRUE, show.aic = TRUE)
```

A multiple logistic regression was calculated to predict the age of the emperors at the end of their reign (age_reign_end).
In MODEL1 rise_recoded and the century was entered as predictive variables.
MODEL1 was significant (F(3, 40) = 69.64, p < 0.05, R^2 = 0.21, RSE = 15.56, RMSE =). The emperor's predicted age_reign_end is equal to 71.561 -15.764 * (rise_recoded) - 5.842 * (century), where rise_recoded is coded as Birthright and Other and century is counted in centuries. Both predictors were significant.
In MODEL2 rise_recoded and age_reign_start was entered as predictive variables.
MODEL2 was significant (F(2, 41) = 105.8, p < 0.001, R^2 = 0.83, RSE = 7.06). The emperor's predicted age_reign_end is equal to -1.96206 + 13.40431 * (rise_recoded) + 1.08973 * (age_reign_start), where rise_recoded is coded as Birthright and Other and age_reign_start is counted in years. Both predictors were significant and were homoscedastic (MODEL1: X2(1, N = 44) = 0.6673762, p > 0.05, MODEL2: X2(1, N = 44) = 1.182378, p > 0.05)
The R^2 is higher and RSE is smaller for MODEL2, which means MODEL2 explains higher variability (83% instead of 21%) of the dependent variable.  According to the visualizations as well, MODEL2 is a better fit to the dataset.

## Probabilities of dying by force, battle, or natural causes

``` {r}
#null model for modeling the predictors of dying by force
model_forced_null <- glm(died_by_force ~ 1, data = emperors_principate, family = binomial)
summary(model_forced_null)
#MODEL3
model_forced <- glm(died_by_force ~ rise_recoded + century + age_reign_start + reign_long_year, data = emperors_principate, family = binomial, control=glm.control(maxit=50))
summary(model_forced)

tab_model(model_forced_null, model_forced, show.est = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE)

#null model for modeling the predictors of dying in battle
model_battle_null <- glm(died_by_battle ~ 1, data = emperors_principate, family = binomial)
summary(model_battle_null)
#MODEL4
model_battle <- glm(died_by_battle ~ rise_recoded + age_reign_start, data = emperors_principate, family = binomial, control=glm.control(maxit=50))
summary(model_battle)

tab_model(model_battle_null, model_battle, show.est = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE)

#null model for modeling the predictors of dying naturally
model_natural_null <- glm(died_naturally ~ 1, data = emperors_principate, family = binomial)
summary(model_natural_null)
#MODEL5
model_natural <- glm(died_naturally ~ rise_recoded + age_reign_start, data = emperors_principate, family = binomial, control=glm.control(maxit=50))
summary(model_natural)

tab_model(model_natural_null, model_natural, show.est = TRUE, show.fstat = TRUE, show.aic = TRUE, show.loglik = TRUE)
```

The predictors were only significant in the case of the predictive model of dying naturally, and the AIC numbers were smaller for the fitted mosel as well (fitted model: 52.224, null model: 55.413). 
```{r}
#creating a tibble with the predicted and observed values and the predictive variables
natural_prob <- predict(model_natural, data = emperors_principate, type = "response")
natural_odds <- natural_prob / (1 - natural_prob)
tibble(name = emperors_principate$name, rise = emperors_principate$rise_recoded, age_reign_start = emperors_principate$age_reign_start, probability = natural_prob, odds = natural_odds, observed = emperors_principate$died_naturally)

#creating a classification table to check the predictions
classDF <- data.frame(response = emperors_principate$died_naturally, predicted = round(fitted(model_natural),0))
class_table <- xtabs(~ predicted + response, data = classDF)
# checking the sensitivity of the model
sensitivity <- class_table[2,2] / (class_table[2,1] + class_table[2,2])
cat("The sensitivity of the model: ", sensitivity)
#checking the specificity of the model
specificity <- class_table[1,1] / (class_table[1,2] + class_table[1,1])
cat("The specificity of the model: ", specificity)

#checking the model with 
res <- chisq.test(emperors_principate$died_naturally, round(fitted(model_natural),0))
res
```


A multiple logistic regression was calculated to predict died_naturally (MODEL5), died_by_force (MODEL3), died_by_battle (MODEL4). The AIC numbers of MODEL3 and MODEL4 were higher then their null models, which means these variables cannot be explained by the variables in the dataset.
The prediction of died_naturally was based on rise_recoded and age_reign_start. The emperor's predicted died_naturally is equal to - 5.24133 + 2.42209 * (rise_recoded)  + 0.07210 * (age_reign_start), where rise_recoded is coded as Birthright and Other and age_reign_start is measured in years. Both predictors were significant.
Based on statistical diagnostics the fitted model cannot be considered reliable. Tjur's R2 is 0.15 so the model does not explain the 85% of the variation,  and the predicted and the observed results was found to be independent (X2(1, N = 44) = 7.2196, p < 0.05).

## Improvements

It would be useful to collect more information related to the emperors: health condition, mental illnesses, foreign policy situation (for eg, war with barbarian tribes), domestic political situation to have a better predictive model.